---
title: "Who wrote this digit? Writer identification with CNNs and Siamese networks"   # working title – change freely
subtitle: "STA5073Z Data Science for Industry 2026 · Assignment 2"
author:
  - name: "Maxine Senderayi (SNDMAX001)"   # Role B – CNN (also GitHub repo and website)
  - name: "Jesse MacDonald (MCDJES006)"    # Role A – data, evaluation design and comparison
  - name: "Nicole Thomas (THMNIC036)"      # Role C – Siamese network
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-align: center
execute:
  echo: true
  warning: false
  message: false

# Shows the coloured "Owner / what goes here" boxes.
# Set to false before submitting – every box disappears from the rendered report.
show-guidance: true
---

```{r}
#| label: setup
#| include: false
# OWNER: Jesse (A). Shared set-up – tell the group before changing anything here.
library(tidyverse)
set.seed(2026)
theme_set(theme_minimal(base_size = 11))

# TRUE once every listed file exists. Use it on any chunk that needs a file:
#   #| eval: !expr ready("outputs/cnn_test_predictions.csv")
# so the report still renders before that file has been produced.
ready <- function(...) all(file.exists(c(...)))

if (ready("data/handwriting.rds")) {
  hw   <- readRDS("data/handwriting.rds")
  meta <- hw$metadata
}
```

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="How to use this file (hidden when show-guidance: false)"}
**Colour = owner.** Green = Jesse (Role A, data, splits and comparison) · Blue = Maxine (Role B, CNN) · Red = Nicole (Role C, Siamese network) · Orange = not assigned yet (decide after Tasks 1–6).

| Report section | Task | Owner | Words | Reads |
|---|---|---|---|---|
| Introduction | 8 | Not assigned yet | ≈ 400 | – |
| Data and exploration | 1 | Jesse (A) | ≈ 550 | `data/handwriting.rds` |
| Evaluation design | 4 | Jesse (A) | ≈ 550 | `data/splits.csv` |
| Writer identification with a CNN | 2 | Maxine (B) | ≈ 650 | `outputs/cnn_test_predictions.csv`, `outputs/cnn_history.csv` |
| Same-writer matching with a Siamese network | 3 | Nicole (C) | ≈ 650 | `outputs/siamese_test_distances.csv`, `outputs/siamese_history.csv` |
| Hyperparameter sensitivity: CNN | 5 | Maxine (B) | ≈ 300 | `outputs/tuning_cnn.csv` |
| Hyperparameter sensitivity: Siamese | 5 | Nicole (C) | ≈ 300 | `outputs/tuning_siamese.csv` |
| CNN vs Siamese: CNN results by digit and session | 6 | Maxine (B) | ≈ 300 | Maxine's own outputs |
| CNN vs Siamese: Siamese results by digit and session | 6 | Nicole (C) | ≈ 300 | Nicole's own outputs |
| CNN vs Siamese: side by side | 6 | Jesse (A) | ≈ 400 | `outputs/comparison.csv` |
| Conclusion | 8 | Not assigned yet | ≈ 300 | – |
| **Total** | | | **≈ 4,700** | Limit 5,000 (excludes AI use statement, appendices, captions, references) |

**House rules**

- Edit only your own sections. To change someone else's, tell them or open a pull request.
- Write in your own words. The brief allows AI for proofreading only, not first drafts.
- Model training lives in scripts (`R/` or `python/`) that write to `outputs/`. The report only reads those files, so it renders in seconds. Commit the scripts too: that is what makes it reproducible.
- Label figures and tables with your section key: `eda`, `split`, `cnn`, `siam`, `tune`, `comp` (e.g. `fig-cnn-digit`, `tbl-siam-metrics`). Cross-reference with `@fig-cnn-digit` and `@sec-cnn`.
- Numbers in the text come from inline `r` code reading the output files, never typed by hand. They then stay correct when a model is re-run.
- Guard every chunk that needs a file with `#| eval: !expr ready("path")`.
- Keep a note of your AI use per task as you go. It goes in the appendix table at the end.
:::
:::

<!-- =====================================================================
     INTRODUCTION  ·  Task 8  ·  OWNER: not assigned yet  ·  ≈ 400 words
     ===================================================================== -->

# Introduction {#sec-intro}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Owner: not assigned yet · Task 8 · ≈ 400 words"}
**Must cover:** the two questions (which of 20 known writers wrote this image, and did the same person write these two images?) · the data in one paragraph (20 writers × 5 digits × 5 sessions × 50 replicates, simulated) · why writer identification matters in practice · a one-line roadmap of the report.
:::
:::

<!-- =====================================================================
     DATA AND EXPLORATION  ·  Task 1  ·  OWNER: Jesse (A)  ·  ≈ 550 words
     ===================================================================== -->

# Data and exploration {#sec-eda}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-tip title="Owner: Jesse (Role A) · Task 1 · ≈ 550 words"}
**The brief asks for:** examples across people, digits and sessions · the visible sources of variation · the limitations of this simulated data.

**Answer these:**

- What differs between writers? Between sessions of the same writer? Between replicates within one session?
- Which simple features vary most (ink amount, width and height, centre of mass, slant)? Do they vary more by writer or by session?
- How much variance sits between writers, between sessions and within a session (variance decomposition)?
- Do writers separate in a PCA of the raw pixels? Do sessions?
- What makes this data easier or less realistic than real handwriting (simulated, digits 0–4 only, 20 writers, perfectly balanced, clean and centred images)?

**Exhibits (labels to use):** `fig-eda-grid` writers × digits · `fig-eda-sessions` one writer and digit across sessions 1–5 · `fig-eda-means` mean and difference images · `fig-eda-features` feature boxplots · `tbl-eda-variance` variance decomposition · `fig-eda-pca` PCA coloured by writer and session.

**Reads:** `data/handwriting.rds`. EDA is fast enough to run inside this file.
:::
:::

```{r}
#| label: fig-eda-grid
#| fig-cap: "One image per writer (rows) and digit (columns), session 1, replicate 1."
#| fig-height: 10
#| eval: !expr ready("data/handwriting.rds")
# Jesse: grid of example images across writers and digits
show_img <- function(m, col = gray.colors(256, start = 1, end = 0), zlim = c(0, 255), ...) {
  image(t(m)[, nrow(m):1], col = col, axes = FALSE, asp = 1, zlim = zlim, ...)
}

par(mfrow = c(20, 5), mar = c(0.1, 0.1, 0.1, 0.1), oma = c(1, 3, 2, 1))
for (w in 1:20) {
  for (d in 0:4) {
    idx <- which(meta$person_id == w & meta$digit == d & meta$session == 1 & meta$replicate == 1)
    show_img(hw$images[idx, , , 1])
    if (w == 1) mtext(d, side = 3, line = 0.2, cex = 0.7)
    if (d == 0) mtext(w, side = 2, line = 0.2, cex = 0.6, las = 1)
  }
}
mtext("digit", side = 3, outer = TRUE, line = 0.7, cex = 0.9)
mtext("writer", side = 2, outer = TRUE, line = 1.7, cex = 0.9)
```

```{r}
#| label: fig-eda-sessions
#| fig-cap: "Writer 1's digit 0, four replicates (rows) across sessions 1–5 (columns). Later sessions show a visibly heavier, more angular stroke."
#| eval: !expr ready("data/handwriting.rds")
# Jesse: same writer and digit across the five sessions, a few replicates each
w0 <- 1; d0 <- 0
par(mfrow = c(4, 5), mar = c(0.1, 0.1, 0.1, 0.1), oma = c(1, 3, 2, 1))
for (r in 1:4) {
  for (s in 1:5) {
    idx <- which(meta$person_id == w0 & meta$digit == d0 & meta$session == s & meta$replicate == r)
    show_img(hw$images[idx, , , 1])
    if (r == 1) mtext(paste("s", s), side = 3, line = 0.2, cex = 0.7)
    if (s == 1) mtext(paste("r", r), side = 2, line = 0.2, cex = 0.7)
  }
}
```

```{r}
#| label: fig-eda-means
#| fig-cap: "Top: mean digit-0 image per writer. Bottom: four writers' session means minus their own overall mean (blue = less ink, red = more ink than that writer's average), showing how each writer's stroke shifts across sessions."
#| fig-height: 9
#| eval: !expr ready("data/handwriting.rds")
# Jesse: mean and difference images
div_col <- colorRampPalette(c("steelblue", "white", "firebrick"))(256)
mean_image <- function(idx) apply(hw$images[idx, , , 1], c(2, 3), mean)

par(mfrow = c(8, 5), mar = c(0.1, 0.1, 0.6, 0.1), oma = c(1, 3, 1, 1))

# writer means, digit 0 (restricting to one digit avoids blurring different digit shapes together)
for (w in 1:20) {
  idx <- which(meta$person_id == w & meta$digit == 0)
  show_img(mean_image(idx))
  mtext(w, side = 3, line = 0.1, cex = 0.55)
}

# session-drift diff images for 4 representative writers x 5 sessions, digit 0
sel_writers <- c(1, 6, 13, 18)
diffs <- list()
for (w in sel_writers) {
  idx_w <- which(meta$person_id == w & meta$digit == 0)
  writer_mean <- mean_image(idx_w)
  for (s in 1:5) {
    idx_s <- which(meta$person_id == w & meta$digit == 0 & meta$session == s)
    diffs[[paste(w, s)]] <- mean_image(idx_s) - writer_mean
  }
}
zmax <- max(abs(unlist(diffs)))
for (w in sel_writers) {
  for (s in 1:5) {
    show_img(diffs[[paste(w, s)]], col = div_col, zlim = c(-zmax, zmax))
    if (w == sel_writers[1]) mtext(paste("s", s), side = 3, line = 0.1, cex = 0.6)
    if (s == 1) mtext(paste("w", w), side = 2, line = 0.1, cex = 0.6)
  }
}
```

```{r}
#| label: fig-eda-features
#| fig-cap: "Five simple image features, boxplotted by writer, session and digit. Ink, width, height and slant separate clearly by writer; none separate visibly by session; digit mostly affects ink, width and height (shape), as expected."
#| fig-height: 8
#| eval: !expr ready("data/handwriting.rds")
# Jesse: ink amount, width/height, centre of mass, slant
rows_idx <- 1:28; cols_idx <- 1:28
n_img <- dim(hw$images)[1]
feat <- tibble(
  image_id = meta$image_id,
  ink = NA_real_, width = NA_real_, height = NA_real_,
  com_x = NA_real_, com_y = NA_real_, slant = NA_real_
)
for (i in seq_len(n_img)) {
  m <- hw$images[i, , , 1]
  ink <- sum(m)
  rs <- which(rowSums(m) > 0); cs <- which(colSums(m) > 0)
  feat$ink[i] <- ink
  feat$width[i] <- if (length(cs)) max(cs) - min(cs) + 1 else NA
  feat$height[i] <- if (length(rs)) max(rs) - min(rs) + 1 else NA
  cx <- sum(m * matrix(cols_idx, 28, 28, byrow = TRUE)) / ink
  cy <- sum(m * matrix(rows_idx, 28, 28)) / ink
  feat$com_x[i] <- cx
  feat$com_y[i] <- cy
  mu11 <- sum(m * outer(rows_idx - cy, cols_idx - cx)) / ink
  mu02 <- sum(m * (rows_idx - cy)^2) / ink
  feat$slant[i] <- if (mu02 > 0) mu11 / mu02 else NA
}
feat <- feat %>%
  mutate(centroid_shift = sqrt((com_x - 14.5)^2 + (com_y - 14.5)^2)) %>%
  select(image_id, ink, width, height, centroid_shift, slant) %>%
  left_join(meta, by = "image_id")

level_order <- as.character(sort(unique(c(feat$person_id, feat$session, feat$digit))))
feat_long <- feat %>%
  pivot_longer(c(ink, width, height, centroid_shift, slant), names_to = "feature", values_to = "value")
plot_df <- bind_rows(
  feat_long %>% transmute(feature, value, group_var = "writer", group_value = person_id),
  feat_long %>% transmute(feature, value, group_var = "session", group_value = session),
  feat_long %>% transmute(feature, value, group_var = "digit", group_value = digit)
) %>%
  mutate(
    group_value = factor(group_value, levels = level_order),
    feature = factor(feature, levels = c("ink", "width", "height", "centroid_shift", "slant")),
    group_var = factor(group_var, levels = c("writer", "session", "digit"))
  )

ggplot(plot_df, aes(group_value, value)) +
  geom_boxplot(outlier.size = 0.3, linewidth = 0.3) +
  facet_grid(feature ~ group_var, scales = "free") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6)) +
  labs(x = NULL, y = NULL)
```

```{r}
#| label: tbl-eda-variance
#| tbl-cap: "Share of feature variance (after removing each feature's digit mean) between writers, between sessions within a writer, and within a session."
#| eval: !expr ready("data/handwriting.rds")
# Jesse: variance decomposition of the features above
# Each feature is first centred on its own digit mean, so digit-shape
# differences (which dominate raw ink/width/height) don't swamp the
# writer/session decomposition we actually care about here.
feat_c <- feat %>%
  group_by(digit) %>%
  mutate(across(c(ink, width, height, centroid_shift, slant), ~ . - mean(.), .names = "resid_{.col}")) %>%
  ungroup()

decompose_one <- function(resid_col) {
  d <- feat_c %>% transmute(resid = .data[[resid_col]], person_id = factor(person_id), session = factor(session))
  m <- aov(resid ~ person_id / session, data = d)
  ss <- summary(m)[[1]][["Sum Sq"]]
  tibble(
    `Between writer %` = round(ss[1] / sum(ss) * 100, 1),
    `Between session %` = round(ss[2] / sum(ss) * 100, 1),
    `Within session %` = round(ss[3] / sum(ss) * 100, 1)
  )
}
feature_cols <- c("resid_ink", "resid_width", "resid_height", "resid_centroid_shift", "resid_slant")
var_tbl <- map_dfr(feature_cols, decompose_one) %>%
  mutate(Feature = c("ink", "width", "height", "centroid_shift", "slant"), .before = 1)
knitr::kable(var_tbl)
```

```{r}
#| label: fig-eda-pca
#| fig-cap: "PCA of raw pixels. Top: a 5,000-image sample across all digits — clusters correspond to digit shape, not writer, and writers don't visibly separate once digit is mixed in. Bottom: restricting to digit 0 only (removing the digit-shape confound) reveals tight, clearly separated per-writer clusters, which are not further split by session — writer style is a strong pixel-level signal once digit is held fixed."
#| fig-height: 9
#| eval: !expr ready("data/handwriting.rds")
# Jesse: PCA of pixels
library(patchwork) # only used in this chunk; not added to shared setup
set.seed(2026)
idx_all <- sample(nrow(meta), 5000)
X_all <- matrix(hw$images[idx_all, , , 1], nrow = length(idx_all))
pc_all <- prcomp(X_all, center = TRUE, scale. = FALSE, rank. = 10)
ve_all <- round(100 * pc_all$sdev[1:2]^2 / sum(pc_all$sdev^2), 1)

idx_d0 <- which(meta$digit == 0)
X_d0 <- matrix(hw$images[idx_d0, , , 1], nrow = length(idx_d0))
pc_d0 <- prcomp(X_d0, center = TRUE, scale. = FALSE, rank. = 10)
ve_d0 <- round(100 * pc_d0$sdev[1:2]^2 / sum(pc_d0$sdev^2), 1)

df_all <- meta[idx_all, ] %>% mutate(PC1 = pc_all$x[, 1], PC2 = pc_all$x[, 2])
df_d0  <- meta[idx_d0, ]  %>% mutate(PC1 = pc_d0$x[, 1],  PC2 = pc_d0$x[, 2])

p1 <- ggplot(df_all, aes(PC1, PC2, color = factor(digit))) +
  geom_point(size = 0.6, alpha = 0.5) +
  labs(color = "digit", title = "All digits, coloured by digit",
       x = paste0("PC1 (", ve_all[1], "%)"), y = paste0("PC2 (", ve_all[2], "%)")) +
  theme_minimal(base_size = 9)

p2 <- ggplot(df_all, aes(PC1, PC2, color = factor(person_id))) +
  geom_point(size = 0.6, alpha = 0.5) +
  labs(color = "writer", title = "All digits, coloured by writer",
       x = paste0("PC1 (", ve_all[1], "%)"), y = paste0("PC2 (", ve_all[2], "%)")) +
  theme_minimal(base_size = 9) + theme(legend.position = "none")

p3 <- ggplot(df_d0, aes(PC1, PC2, color = factor(person_id))) +
  geom_point(size = 0.8, alpha = 0.6) +
  labs(color = "writer", title = "Digit 0 only, coloured by writer",
       x = paste0("PC1 (", ve_d0[1], "%)"), y = paste0("PC2 (", ve_d0[2], "%)")) +
  theme_minimal(base_size = 9) + theme(legend.position = "none")

p4 <- ggplot(df_d0, aes(PC1, PC2, color = factor(session))) +
  geom_point(size = 0.8, alpha = 0.6) +
  labs(color = "session", title = "Digit 0 only, coloured by session",
       x = paste0("PC1 (", ve_d0[1], "%)"), y = paste0("PC2 (", ve_d0[2], "%)")) +
  theme_minimal(base_size = 9)

(p1 + p2) / (p3 + p4)
```

<!-- =====================================================================
     EVALUATION DESIGN  ·  Task 4  ·  OWNER: Jesse (A)  ·  ≈ 550 words
     ===================================================================== -->

# Evaluation design {#sec-split}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-tip title="Owner: Jesse (Role A) · Task 4 · ≈ 550 words"}
**The brief asks for:** the chosen train, validation and test strategy and why · what kind of generalisation the evaluation represents · possible leakage or optimistic performance.

**Answer these:**

- Main split (both models): sessions 1–3 train, 4 validation, 5 test. What does it test? (A known writer on a new day.)
- Unseen-writer split (Siamese only): 14 / 3 / 3 writers. What does it test? (A person the model has never seen.) Why can the CNN not do this?
- What is validation used for (early stopping, tuning, the Siamese threshold)? Why is test touched only once?
- Optimism experiment: how much better does a simple baseline look on a random image split than on the session-held-out split?
- Leakage risks: near-duplicate replicates within a session, pairs built before splitting, a threshold tuned on test, the same image in training and test pairs.
- Why might results still be optimistic even with a clean split (simulated data, 20 writers)?

**Exhibits:** `tbl-split-counts` images per split · `tbl-split-optimism` (or `fig-split-optimism`) baseline accuracy on the two splits.

**Reads:** `data/splits.csv`. **Code:** `R/00_data.R` writes the split; `R/optimism_baseline.R` runs the experiment.
:::
:::

```{r}
#| label: tbl-split-counts
#| tbl-cap: "Number of images in each split. split_main is the session-based split used by both models (sessions 1-3/4/5 = train/val/test); split_writer is the writer-held-out split (14/3/3 writers = train/val/test), used only for the Siamese network's unseen-writer experiment."
#| eval: !expr ready("data/splits.csv")
splits <- read_csv("data/splits.csv", show_col_types = FALSE)
# Jesse: counts by split_main and split_writer
bind_rows(
  splits %>% count(split = split_main) %>% mutate(strategy = "split_main (session-based)"),
  splits %>% count(split = split_writer) %>% mutate(strategy = "split_writer (writer-based)")
) %>%
  mutate(split = factor(split, levels = c("train", "val", "test"))) %>%
  pivot_wider(names_from = split, values_from = n) %>%
  knitr::kable()
```

```{r}
#| label: tbl-split-optimism
#| tbl-cap: "Baseline writer accuracy on a random image split vs the session-held-out split."
#| eval: !expr ready("outputs/optimism_baseline.csv")
# Jesse: read outputs/optimism_baseline.csv and show the gap
```

<!-- =====================================================================
     CNN  ·  Task 2  ·  OWNER: Maxine (Role B)  ·  ≈ 650 words
     ===================================================================== -->

# Writer identification with a CNN {#sec-cnn}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-note title="Owner: Maxine (Role B) · Task 2 · ≈ 650 words"}
**The brief asks for:** a prediction task whose response is writer ID · a CNN, built and evaluated · overall writer identification performance · performance compared by digit · what the evaluation measures mean in real-world terms.

**Answer these:**

- What exactly is predicted? (Input: one 28 × 28 image. Output: one of 20 known writers, a closed set.) Point to @sec-split for the split rather than repeating it.
- Architecture and training: layers, embedding size, optimiser, learning rate, batch size, epochs, early stopping rule.
- Overall test results: accuracy, top-3 accuracy, macro-F1. Which writers get confused with each other?
- Accuracy by digit, with 95% intervals (1,000 test images per digit).
- What do the numbers mean in practice? The chance of naming the right writer for a sample written on a new day; top-3 as a shortlist for a human examiner; the model always names one of the 20, even for an unknown writer.

**Exhibits:** `tbl-cnn-arch` layers · `fig-cnn-training` loss and accuracy curves · `tbl-cnn-metrics` overall metrics · `fig-cnn-confusion` confusion matrix · `fig-cnn-digit` accuracy by digit.

**Reads:** `outputs/cnn_test_predictions.csv`, `outputs/cnn_history.csv`. **Code:** `R/cnn.R`.
:::
:::

```{r}
#| label: tbl-cnn-arch
#| tbl-cap: "CNN architecture."
#| eval: false
# Maxine: layer table (can be typed by hand or taken from summary(model))
```

```{r}
#| label: fig-cnn-training
#| fig-cap: "CNN training and validation loss and accuracy by epoch."
#| eval: !expr ready("outputs/cnn_history.csv")
# Maxine: training curves
```

```{r}
#| label: tbl-cnn-metrics
#| tbl-cap: "CNN writer identification on the session-5 test set."
#| eval: !expr ready("outputs/cnn_test_predictions.csv")
cnn_pred <- read_csv("outputs/cnn_test_predictions.csv", show_col_types = FALSE)
# Maxine: accuracy, top-3 accuracy, macro-F1
```

```{r}
#| label: fig-cnn-confusion
#| fig-cap: "Confusion matrix of true vs predicted writer on the test set."
#| eval: !expr ready("outputs/cnn_test_predictions.csv")
# Maxine: confusion matrix heatmap
```

```{r}
#| label: fig-cnn-digit
#| fig-cap: "CNN test accuracy by digit, with 95% intervals."
#| eval: !expr ready("outputs/cnn_test_predictions.csv")
# Maxine: accuracy by digit
```

<!-- =====================================================================
     SIAMESE NETWORK  ·  Task 3  ·  OWNER: Nicole (Role C)  ·  ≈ 650 words
     ===================================================================== -->

# Same-writer matching with a Siamese network {#sec-siam}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-important title="Owner: Nicole (Role C) · Task 3 · ≈ 650 words"}
**The brief asks for:** a Siamese network that estimates whether two images come from the same writer · contrastive or triplet loss · how training examples are constructed · how matching performance is evaluated.

**Answer these:**

- Architecture: shared encoder (same backbone as the CNN), embedding size, distance.
- Loss: contrastive or triplet, the margin, and why.
- Training examples: how positive and negative pairs are sampled, why digit is held fixed, the same/different balance, cross-session positives, and pairs built only after splitting.
- Threshold: chosen on validation pairs by balanced accuracy.
- Test results on `data/test_pairs.csv`: ROC AUC, equal error rate, balanced accuracy, and distance histograms for same vs different pairs.
- Unseen-writer experiment: the same metrics on the 3 held-out writers.
- What the errors mean: a false accept matches two different writers; a false reject misses a true match. Which matters more depends on the use.

**Exhibits:** `fig-siam-pairs` example positive and negative pairs · `fig-siam-dist` distance histograms · `fig-siam-roc` ROC curves · `tbl-siam-metrics` main split vs unseen writers.

**Reads:** `outputs/siamese_test_distances.csv`, `outputs/siamese_history.csv`. **Code:** `R/siamese.R` (or `python/siamese.py`).
:::
:::

```{r}
#| label: fig-siam-pairs
#| fig-cap: "Example positive (same writer) and negative (different writer) training pairs."
#| eval: !expr ready("data/handwriting.rds", "data/test_pairs.csv")
# Nicole: a few pairs of each kind
```

```{r}
#| label: fig-siam-dist
#| fig-cap: "Embedding distances for same-writer and different-writer test pairs."
#| eval: !expr ready("outputs/siamese_test_distances.csv")
siam_dist <- read_csv("outputs/siamese_test_distances.csv", show_col_types = FALSE)
# Nicole: histograms, with the validation threshold marked
```

```{r}
#| label: fig-siam-roc
#| fig-cap: "ROC curves for same-writer matching."
#| eval: !expr ready("outputs/siamese_test_distances.csv")
# Nicole: ROC curve(s)
```

```{r}
#| label: tbl-siam-metrics
#| tbl-cap: "Siamese matching performance: known writers on a new session vs unseen writers."
#| eval: !expr ready("outputs/siamese_test_distances.csv")
# Nicole: AUC, EER, balanced accuracy for both settings
```

<!-- =====================================================================
     HYPERPARAMETER SENSITIVITY  ·  Task 5  ·  OWNERS: Maxine (CNN) and Nicole (Siamese)
     ===================================================================== -->

# Hyperparameter sensitivity {#sec-tune}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Shared section · Task 5"}
**The brief asks for:** a limited hyperparameter search for at least one model (we do both) · how sensitive the evaluation measures are to the hyperparameters varied. Maxine writes the CNN part and Nicole the Siamese part. Use the same plot style so the two figures read alike.
:::
:::

<!-- ---- CNN tuning · OWNER: Maxine (Role B) · ≈ 300 words ---- -->

## CNN {#sec-tune-cnn}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-note title="Owner: Maxine (Role B) · Task 5 · ≈ 300 words"}
**Answer these:**

- Which 2–3 hyperparameters were varied, over which values, and why those?
- Design: grid or random search, number of configurations, seeds per configuration. Chosen on validation only.
- How much does validation accuracy move across each hyperparameter, compared with seed-to-seed noise?
- The chosen configuration and its single test result.

**Exhibits:** `fig-tune-cnn`. **Reads:** `outputs/tuning_cnn.csv`. **Code:** `R/tune_cnn.R`.
:::
:::

```{r}
#| label: fig-tune-cnn
#| fig-cap: "CNN validation accuracy across hyperparameter values (points = seeds)."
#| eval: !expr ready("outputs/tuning_cnn.csv")
tune_cnn <- read_csv("outputs/tuning_cnn.csv", show_col_types = FALSE)
# Maxine: sensitivity plot
```

<!-- ---- Siamese tuning · OWNER: Nicole (Role C) · ≈ 300 words ---- -->

## Siamese network {#sec-tune-siam}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-important title="Owner: Nicole (Role C) · Task 5 · ≈ 300 words"}
**Answer these:**

- Which hyperparameters were varied (e.g. margin, embedding size, learning rate, pairs per epoch), over which values, and why?
- Design: number of configurations and seeds. Chosen on validation AUC only.
- How much does validation AUC move across each hyperparameter, compared with seed-to-seed noise?
- The chosen configuration and its single test result.

**Exhibits:** `fig-tune-siam`. **Reads:** `outputs/tuning_siamese.csv`. **Code:** `R/tune_siamese.R`.
:::
:::

```{r}
#| label: fig-tune-siam
#| fig-cap: "Siamese validation AUC across hyperparameter values (points = seeds)."
#| eval: !expr ready("outputs/tuning_siamese.csv")
tune_siam <- read_csv("outputs/tuning_siamese.csv", show_col_types = FALSE)
# Nicole: sensitivity plot, same style as fig-tune-cnn
```

<!-- =====================================================================
     CNN vs SIAMESE  ·  Task 6  ·  OWNERS: Maxine, Nicole, Jesse
     ===================================================================== -->

# CNN vs Siamese network: digits and sessions {#sec-comp}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Shared section · Task 6"}
**The brief asks for:** a comparison of the CNN and Siamese results · a critical discussion of which digits carry more writer information and how session changes affect performance. Maxine and Nicole each report their own model's digit and session results; Jesse then compares the two side by side.
:::
:::

<!-- ---- CNN by digit and session · OWNER: Maxine (Role B) · ≈ 300 words ---- -->

## CNN results by digit and session {#sec-comp-cnn}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-note title="Owner: Maxine (Role B) · Task 6 · ≈ 300 words"}
**Answer these:**

- Digit-only CNNs (one model per digit): which digit alone identifies writers best and worst?
- Does that ranking match the per-digit accuracy of the full model in @sec-cnn?
- Leave-one-session-out: accuracy for each held-out session. Is any session unusual?

**Exhibits:** `fig-comp-cnn-digitonly` · `fig-comp-cnn-loso`. **Code:** `R/cnn_digit_session.R`.
:::
:::

```{r}
#| label: fig-comp-cnn-digitonly
#| fig-cap: "Test accuracy of CNNs trained on a single digit."
#| eval: !expr ready("outputs/cnn_digit_only.csv")
# Maxine: accuracy by digit for digit-only models
```

```{r}
#| label: fig-comp-cnn-loso
#| fig-cap: "CNN accuracy for each held-out session (leave-one-session-out)."
#| eval: !expr ready("outputs/cnn_loso.csv")
# Maxine: accuracy by held-out session
```

<!-- ---- Siamese by digit and session · OWNER: Nicole (Role C) · ≈ 300 words ---- -->

## Siamese results by digit and session {#sec-comp-siam}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-important title="Owner: Nicole (Role C) · Task 6 · ≈ 300 words"}
**Answer these:**

- AUC by digit: which digits separate writers best and worst?
- Same-session vs cross-session positive pairs: how much larger are distances across sessions?
- Cross-digit pairs: does writer identity carry across digits?

**Exhibits:** `fig-comp-siam-digit` · `fig-comp-siam-session`. **Reads:** `outputs/siamese_test_distances.csv`.
:::
:::

```{r}
#| label: fig-comp-siam-digit
#| fig-cap: "Siamese matching AUC by digit."
#| eval: !expr ready("outputs/siamese_test_distances.csv")
# Nicole: AUC by digit
```

```{r}
#| label: fig-comp-siam-session
#| fig-cap: "Distances for same-writer pairs from the same session vs different sessions."
#| eval: !expr ready("outputs/siamese_test_distances.csv")
# Nicole: same- vs cross-session distances
```

<!-- ---- Side by side · OWNER: Jesse (A) · ≈ 400 words ---- -->

## Side by side {#sec-comp-both}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-tip title="Owner: Jesse (Role A) · Task 6 · ≈ 400 words"}
**Answer these:**

- With one script on the same test pairs, how do the PCA baseline, CNN and Siamese embeddings compare on AUC, equal error rate and nearest-centroid writer accuracy?
- Which model suits which question (identify one of 20 known writers vs verify a pair), and what does each cost (e.g. retraining when a new writer is added)?
- Do the digit rankings agree across the CNN, the Siamese network and the EDA?
- How large is the session gap (same-session vs cross-session) for each model?
- Critical caveats: simulated data, one test session, 20 writers.

**Exhibits:** `tbl-comp-summary` · `fig-comp-digit` AUC by digit for PCA, CNN and Siamese.

**Reads:** `outputs/comparison.csv`. **Code:** `R/eval_embeddings.R` (reads `outputs/cnn_embeddings.csv`, `outputs/siamese_embeddings.csv`, `data/test_pairs.csv`).
:::
:::

```{r}
#| label: tbl-comp-summary
#| tbl-cap: "PCA baseline, CNN and Siamese embeddings scored on the same test pairs."
#| eval: !expr ready("outputs/comparison.csv")
comparison <- read_csv("outputs/comparison.csv", show_col_types = FALSE)
# Jesse: AUC, EER, nearest-centroid accuracy; same- vs cross-session
```

```{r}
#| label: fig-comp-digit
#| fig-cap: "Matching AUC by digit for the PCA baseline, CNN and Siamese network."
#| eval: !expr ready("outputs/comparison.csv")
# Jesse: AUC by digit, one line or bar group per model
```

<!-- =====================================================================
     CONCLUSION  ·  Task 8  ·  OWNER: not assigned yet  ·  ≈ 300 words
     ===================================================================== -->

# Conclusion {#sec-conclusion}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Owner: not assigned yet · Task 8 · ≈ 300 words"}
**Must cover:** the answer to each question in one or two sentences · the main limitation · what you would do next with more time or real data.
:::
:::

# References {.unnumbered}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Owner: not assigned yet"}
Add a `references.bib` file and `bibliography: references.bib` to the YAML header, then cite with `[@key]`. The list appears here automatically.
:::
:::

::: {#refs}
:::

<!-- =====================================================================
     APPENDICES  ·  not counted in the 5,000 words
     ===================================================================== -->

# Plagiarism declaration {.appendix}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Owner: not assigned yet"}
Paste the course's standard plagiarism declaration here, with all three names.
:::
:::

# AI use statement {.appendix}

::: {.content-visible when-meta="show-guidance"}
::: {.callout-warning title="Each person fills the rows for their own tasks"}
The brief asks, for each task: (a) what tool you used, (b) what you used it for, (c) if and how you checked its outputs, and (d) how much of what it produced you ended up using.
:::
:::

| Task | Filled by | (a) Tool | (b) Used for | (c) How checked | (d) How much used |
|---|---|---|---|---|---|
| 1 Data exploration | Jesse | | | | |
| 2 CNN | Maxine | | | | |
| 3 Siamese network | Nicole | | | | |
| 4 Evaluation design | Jesse | | | | |
| 5 Hyperparameter sensitivity | Maxine, Nicole | | | | |
| 6 Comparison | Jesse, Maxine, Nicole | | | | |
| 7 Git and GitHub | Maxine | | | | |
| 8 Report | All | | | | |
| 9 GitHub Pages | Maxine | | | | |
